\plaintitle

### 2025-03-13 docs/style/article-body.tex
added one related work

In this project, we will be writing firmware for one such solution, which reads quantum variations from an optical signal. Further details about how this signal is produced will be introduced in section 2 and builds on the work of Clason \cite{Clason2023}. This optical signal will be converted to a stream of random, raw bits via an Analog to Digital Converter (\emph{ADC}). In turn, these random bits will be processed via Toeplitz extraction \cite{toeplitz} in order to process these bits into random numbers. Some processing has to be done on the microcontrollers themselves in order to ensure that the data is workable, and Toeplitz extraction is a tried and tested method to accomplish this. These random numbers will then be output from the microcontroller to the host computer via USB. This thesis will aim to answer one key research question: How can sampled vacuum fluctuations be processed efficiently in order to output QRNG?

@@

This analog current isn't suitable to operate on without further processing. As mentioned in section 1, the signal needs to pass through an ADC and converted into raw bits in order for it to be usable. In his thesis, Clason \cite{Clason2023} suggests a discrete ADC chip capable of analyzing frequencies higher than 25 MHz, as this is the highest frequency studied in his work. The Nyquist-Shannon theorem poses that, in order to accurately reconstruct the signal, the sampling rate must be at least twice the highest frequency component present in the signal, \(f_s \geq 2 f_{\text{max}}\). \textbf{CITATION NEEDED}

\textbf{This section will be filled with more information as soon as the exact ADC we will be using has been selected. The exact requirements needs to be discussed with the project owner before a choice can be made!}

@@

Microcontrollers (MCUs) are compact and low-power computing devices designed primary for embedded systems and real-time applications. In contrast to general purpose CPUs, MCUs integrates a processor, memory, and peripherals into a single chip, thus enabling efficient and autonomous operation in constrained environments. To program MCUs, various development frameworks and tools such as: Arduino Framework, Espressif IDF, raw C++, etc. Different frameworks can contribute to overhead.

In our project, we will be utilizing several microcontrollers. Primarily, we will be using Teensy 4.1\footnote{\href{https://www.pjrc.com/store/teensy41.html}{Teensy developer documentation, accessed 2025-02-27}} as our baseline, as this controller is relatively cheap yet very powerful.

@@

\textbf{WIP}

@@

With the assumption that the OQRNG-device produces a truly random analog signal, we can now clearly define the scope in which this thesis operates. Considering the maximum conversion speed from the ADC and the USB-output from the MCU, we have a clear bound between \textbf{PLACEHOLDER} and \textbf{PLACEHOLDER} in which Toeplitz extraction needs to be executed. Any implementation of Toeplitz extraction must then execute fast enough on any given microcontroller feasible for the proposed quantum RNG-thumbstick as to not be the decisive limiting factor.

@@

\subsubsection{\texorpdfstring{5.2 Evaluation }{5.2 Evaluation }}\label{evaluation}

Due to the bounds introduced by the hardware limitations, evaluating the throughput of each iteration essentially consists of measuring the execution time. In order to verify accuracy, the output of random numbers should be identical for each sample bitstring tested. Certain iterations might, however, use larger parts of the bitstring, and thus produce a slightly variable result.

As discussed in section 3, the hardware used will impose a clear bound on how quickly our implementation needs to process the bits in order to match the speed of the ADC, as well as the output speed of the USB-port, both in \(MB/s\). Hyncica et. al. \cite{micromeasurements} propose that measuring execution time of algorithms directly via the microcontrollers internal timers (while subtracting the interrupt overhead) provides adequate measurements. An additional advantage is that the same code can be used to measure execution speed on several different microcontrollers, rather than relying on counting CPU cycles (as the process for this may vary greatly between controllers). As we will use fixed-size bitstrings for evaluation, we can then derive the throughput of the algorithm in \(MB/s\) as follows:

\[
Throughput_{MB/s} = \frac{DataSize_{bits}}{Execution
Time_{ms}} \times \frac{1}{8} \times \frac{1000}{10^6} \text{\phantom{12}(1)}
\]

@@

\newpage

\phantom{HAHA I CAN CURSE HERE! FUCK YOU!} \newpage

@@

2025-03-XX: Moved evaluation down in the methodology in order to provide a better flow. Elaborated further on Toeplitz extraction and ADC converters, as well as motivating the selection of these.

>>>>

In this project, we will be writing firmware for a quantum number generator, which generates true randomness by reading shot noise from an optical signal. Further details about how this signal is produced will be introduced in section 2 and 3, and builds on the work of Clason \cite{Clason2023}. This optical signal will be converted to a stream of random, raw bits via an Analog to Digital Converter (\emph{ADC}). In turn, these random bits will be processed via Toeplitz extraction \cite{toeplitz} in order to process these bits into random numbers. Some processing has to be done on the microcontrollers themselves in order to ensure that the data is workable, and Toeplitz extraction is a tried and tested method to accomplish this. These random numbers will then be output from the microcontroller to the host computer via USB. This thesis will aim to answer one key research question: How can sampled vacuum fluctuations be processed efficiently in order to output QRNG?

@@

This analog current isn't suitable to operate on without further processing. As mentioned in section 1, the signal needs to pass through an ADC and converted into raw bits in order for it to be usable. In his thesis, Clason \cite{Clason2023} suggests a discrete ADC chip capable of analyzing frequencies higher than 25 MHz, as this is the highest frequency studied in his work. However, in the interest of keeping the implementation light and cheap, we will be using ADCs that provide less samples per second and lower frequencies. This is done mainly for ease of development and access to this hardware.

Initially, we will utilize MAX11102AUB\footnote{\href{https://www.farnell.com/datasheets/1913106.pdf}{Technical specification for MAX11102AUB, accessed 2025-03-13}} with an effective sample rate of 2 million samples per second (MSPS). This ADC provides a 12 bit sample size, providing roughly 2.86 MB/s of sampled data per second, derived by the following calculation.

\[
Data Rate = \frac{2,000,000_{MSPS} \times 12}{8 \times 1,048,576} \approx 2.86 MB/s \text{\phantom{123} (1)}
\]

Many microcontrollers furthermore come equipped with internal ADCs that can be utilized, and while these provide a lower sample size (often around 1 MSPS), the ease of development may be prudent to utilize for this proof-of-concept. While our initial ADC has a fairly low throughput, this can always be upgraded if it ends up becoming too limiting. The final output from the ADC, whether built into the microcontroller or an external one, will be a stream of raw bits, as the analog signal from the OQRNG-device is processed.

@@

Microcontrollers (MCUs) are compact and low-power computing devices designed primary for embedded systems and real-time applications. Unlike general purpose CPUs, MCUs integrates a processor, memory, and peripherals into a single chip. Thus, enabling efficient and autonomous operation in constrained environments. Various development frameworks and tools are available for programming MCUs, including the Arduino Framework, manufacturer specific SDKs and raw C++, among others. Each framework presents different trade-offs in terms of performance overhead, efficiency and ease of development.

Modern microcontrollers feature advanced microarchitectural elements to enhance processing capabilities. Some of high performance MCUs utilize dual-issue superscalar architectures, enabling them to run several instructions in one cycle. This type of design greatly improves single-threaded capabilities, which makes these MCUs ideal for tasks that need real-time data processing. The use of dual-issue architectures in embedded devices has demonstrated high execution speeds, along with the added benefit of improved energy efficiency. Thus, making them a great option for applications that require low power but high performance. Research conducted by Lozano et al. \cite{dual-issue} specifically explores energy efficient dual-issue processors designs. This study shows that optimizing instruction throughput and power consumption can significantly improve processing efficiency in embedded systems. While energy efficiency is not the primary goal for this study, ensuring that the resulting USB-device does not consume too much power is tangentially relevant for real-world usage.

Since MCUs often function under strict timing requirements, it is critical to have effective ways to access memory and transfer data for processing in real time. High-performance MCUs enhance memory usage in various ways. Some of them use Tightly Coupled Memory (TCM), which gives fast SRAM with specific access routes for important data, avoiding cache misses and guaranteeing consistent performance. Moreover, instruction and data caching techniques, including instruction prefetching and branch prediction, help minimize execution delays in computationally intensive real-time applications. Another important aspect is Direct Memory Access (DMA), which enables data transfer between peripherals such as the ADC and RAM, without CPU intervention. This offloading reduces processing overhead, allowing the MCU to manage fast data transfers effectively. These improvements are especially significant for Toeplitz extraction, where large amount of random data needs to be processed and sent quickly with low delay. Efficient memory management guarantees that randomness extraction can occur rapidly without major slowdowns in computing. Both approaches will be tested during development.

The Teensy 4.1\footnote{\href{https://www.pjrc.com/store/teensy41.html}{Teensy developer documentation, accessed 2025-02-27}}, based on the ARM Cortex-M7 is especially good for computationally demanding tasks involving randomness extraction due to its dual-issue superscalar architecture and DSP capabilities. The floating-point unit (FPU) and SIMD-style DSP instructions imporove how quickly it can perform bitwise and arithmetic tasks, which are crucial for quick Toeplitz extraction. Although, the Cortex-M7 does not have special features for cryptography, its ability to execute multiple instructions quickly, combined with fast memory access and an efficient pipeline design, allows it to effectively manage and process randomness extraction. Moreover, its USB High-Speed interface allows rapid transmission of extracted random data to an external system, ensuring minimal bottlenecks in high-rate randomness generation.

@@

The raw bits from the ADC can potentially have some deterministic patterns, and as such have to be processed somehow in order to remove these patterns. Several methods exist for this purpose, and for our work, we will perform this preprocessing via Toeplitz extraction. The main focus of this study is to implement this extraction algorithm as effectively as possible on resource constrained hardware.

A detailed account of the inner workings of Toeplitz extraction can be found in the work of Chouhan et al. \cite{toeplitz-desc}. This work focuses on implementing Toeplitz extraction on field-programmable gate-arrays (FPGA), but some specific details can be derived from their work. As these authors describe, Toeplitz extraction is a strong contender for our work due to a lower computational complexity than other alternatives, as well as a relatively easy algorithm to use. This extraction utilizes either matrix multiplication or hashing between a pseudo-random seed and the raw data provided from a high-entropy source of randomness -- in our case, the OQRNG-device.

To summarize the theoretical working of Toeplitz extraction (as explained by Chouhan et al. \cite{toeplitz-desc}), the sampled raw bit matrix (\(T\)) are multiplied with a pre-determined seed matrix (\(K\)). The size of the seed is directly dependant on the size of the sampled data, and can be fixed or continually re-sampled as needed. To ensure high levels of entropy, our intuition is that re-sampling the seed from the OQRNG-device continually is prudent. The sample and seed with then be processed with matrix multiplication to remove deterministic patterns, and produce a bitstring that results in our randomly generated number.

@@

With the assumption that the OQRNG-device produces a truly random analog signal, we can now clearly define the scope in which this thesis operates. Considering the maximum conversion speed from the ADC and the USB-output from the MCU, we have a clear bound between 2.86 MS/s (for the ADC) and 480 MB/s (for the USB output) in which Toeplitz extraction needs to be executed. Any speeds over 2.86 MB/s allows us to upgrade the ADC iteratively to continue increasing the output speed. Any implementation of Toeplitz extraction must then execute fast enough on any given microcontroller feasible for the proposed quantum RNG-thumbstick as to not be the decisive limiting factor.

@@

\subsubsection{5.2 Evaluation}\label{evaluation}

As discussed in section 3, the hardware used will impose a clear bound on how quickly our implementation needs to process the bits in order to match the speed of the ADC, as well as the output speed of the USB-port, both in \(MB/s\). In order to verify accuracy, the output of random numbers should be identical for each sample bitstring tested. Certain iterations might, however, use larger parts of the bitstring, and thus produce a slightly variable result.

Hyncica et. al. \cite{micromeasurements} propose that measuring execution time of algorithms directly via the microcontrollers internal timers (while subtracting the interrupt overhead) provides adequate measurements of an algorithms execution speed. An additional advantage is that the same code can be used to measure execution speed on several different microcontrollers, rather than relying on counting CPU cycles (as the process for this may vary greatly between controllers). As we will use fixed-size bitstrings for evaluation, we can then derive the throughput of the algorithm in \(MB/s\) as follows:

\[
Throughput*{MB/s} = \frac{DataSize*{bits}}{Execution Time_{ms}} \times
\frac{1}{8} \times \frac{1000}{10^6} \text{\phantom{12}(2)}
\]

@@

\subsection{6 EXPERIMENTATION}\label{experimentation}

Each iteration results in an implementation written in C, all of which will be supplied as an appendix to this paper. Furthermore, to simplify the implementation process, we will use the Arduino framework to communicate with the hardware. Whereas this introduces some overhead, it drastically reduces the time spent per iteration, and is deemed to be a suitable starting point.

@@

2025-03-10: Moved evaluation down in the methodology in order to provide a better flow. Elaborated further on Toeplitz extraction and ADC converters, as well as motivating the selection of these. Some additional information added in introduction as motivation for the work.

2025-03-12: Elaborated on background, as well as adding more details regarding hardware. Note that the hardware selected is subject to change over time. Further elaborated on related works in optimizing Toeplitz extraction.

### 2025-03-13 docs/style/article-final.tex
added one related work

2025-03-XX: Moved evaluation down in the methodology in order to provide a better flow. Elaborated further on Toeplitz extraction and ADC converters, as well as motivating the selection of these.
% BALANCE COLUMNS
\balance{}

>>>>

2025-03-10: Moved evaluation down in the methodology in order to provide a better flow. Elaborated further on Toeplitz extraction and ADC converters, as well as motivating the selection of these. Some additional information added in introduction as motivation for the work.

2025-03-12: Elaborated on background, as well as adding more details regarding hardware. Note that the hardware selected is subject to change over time. Further elaborated on related works in optimizing Toeplitz extraction.
% BALANCE COLUMNS
\balance{}
